---
title: "Conditional statements and loops"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

<div class="boxobj">
<strong>Learning Objectives</strong>

*  Fill in learning objectives   
</div>

---

# Conditional statements

Decision making is an important part of programming. This can be achieved in R programming using conditional statements such as `if` and `if/else`.

## if

The syntax of an <b>if</b> statement is:
```{r, eval = FALSE}
if (test_expression) {
  do_this
}

x <- 5
if (x > 0) {
  print("positive number")
}
```

## if/else

The syntax of an <b>if/else</b> statement is:
```{r, eval = FALSE}
if (test_expression) {
  do_this
} else {
  do_that
}
```
The else part is optional and is only evaluated if test_expression is FALSE. It is important that the `else` word be in the same line as the closing brace of the `if` statement.

```  {r}
x <- 1
if (x > 0){
  print("positive number")
} else {
  print("negative number")
}
```

Exercise: write a simple if else statement to chck if 5 is an odd number and if it is print "I am odd", otherwise print "I am even".

# Loops

Conceptually, a loop is a way to repeat a sequence of instructions under certain conditions. They allow you to automate parts of your code that are in need of repetition.

The easiest and most frequently used loop in R is a <b>for</b> loop. Here is a demonstration of using loops.

```{r}
year <- c(2015,2016,2017,2018)

for(i in 1:length(year)) {
  print(year[i])
}

for(i in 1:length(year)) {
  print(paste0("the year is ",year[i]))
}
```

Exercises

