---
title: "Bonus project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Introduction

This data set comes from a short-term evolution experiment investigating adaptation of *Pseudomonas aeruginosa* to cystic fibrosis lung-like conditions. Several populations were evolved under four different environmental conditions and then compared to a set of strains isolated from clinical samples (ie. strains that had evolved in the human lung).

You can find the data set [here](https://www.dropbox.com/s/fflcupr2erl4f36/pa_clinical_distance.csv?dl=0) and you can read more about the experimental setup [here](https://doi.org/10.1073/pnas.1721270115).

<br>

# Task 1 - Plot

Use this data set to produce the plot below.

```{r, echo = FALSE}
data <- read.csv("data/pa_clinical_distance.csv")

ggplot(data, aes(x = type, y = log10(dist))) + geom_point() +
  facet_grid(.~env)
```

<br>

# Task 2 - Fancy plot

Modify your plot so that is looks like this:

<br>

```{r, echo = FALSE}
env_key <- c("6scfm" = "SCFM", "7min" = "MIN", "8scfm_mucin" = "SCFM + mucin", "9min_mucin" = "MIN + mucin")
data <- data %>%
  mutate(env_name = env_key[env])
data$env_name <- factor(data$env_name, levels = c("SCFM", "MIN","SCFM + mucin", "MIN + mucin"))


ggplot(data, aes(type, log10(dist), colour = type)) +
	geom_jitter(size = 2, show.legend = F, width = 0.1) +
	facet_grid(~env_name) +
	ylab("log(distance)") +
	scale_x_discrete(labels = c("Epidemic", "Non")) + 
	scale_colour_manual(values = c("orange", "blue")) +
	ggtitle("Comparison of clinical and evolved isolates") +
	stat_summary(geom = "errorbar", fun.y = mean, width = 0.5, colour = "black", aes(ymax =..y.., ymin =..y..)) +
	xlab("Strain type")
```

Your plot need not be identical, but should plot the same data with the same plot features, including:

* changing the treatment names
* changing the order of the env variable
* change the colours
* remove the legend
* add black bars to signify means
* change axes labels
* jitter points

<br>

# Task 3 - Stats

The pattern in the plot above should be fairly clear, but we still need to 'do stats' on it. Use R to create a linear model with distance as the response variable and treatment and strain type as the explanatory variables. Use this model to answer the following questions:

1. Does treatment have a significant effect? 
2. Does strain type have a significant effect?
3. Is there a significant interaction between the two variables?


