---
title: "Revision session"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tibble)
library(tidyr)
```

## Installing packages from different sources.

1. CRAN
You can do this from tools menubar and select install packages.

2. Bioconductor
```{r}
source("https://bioconductor.org/biocLite.R")
biocLite("affy")

```

3. Github

```{r}
library(devtools)
install_github("hadley/dplyr")
```

4. tar.gz packages
Download the R package and use the tools menu bar -> install pacakages to install it.

## Data Types in R

### Vectors

When you want to create vector with more than one element, you should use c() function which means to combine the elements into a vector.

``` {r, }
# Create a vector.
apple <- c('red','green',"yellow")
print(apple)

# Get the class of the vector.
print(class(apple))
```

<font color ="red"><b> Lets try a few functions:</font> </b>
<pre>
length 
toupper 
tolower 
paste0(paste an 'e' at the end of all colors in apple vector)
grep (grep red from the apple vector)
</pre>

<br>

Vectors can also be numeric like:
```{r}
a <- c(5,5,5,6)
b <- c(15,151,15,61)
a+b
a-b
c<-c(a,b)
c
```

### Matrices

A matrix is a two-dimensional rectangular data set. It can be created using a vector input to the matrix function.

```{r,}
# Create a matrix.
m = matrix( c('a','a','b','c','b','a'), nrow = 2, ncol = 3, byrow = TRUE)

print(m)

```

<font color ="red"><b> Lets try a few functions:</font> </b>
<pre>
dim()
ncol()
nrow()
class()
str()
length()
</pre>


### Factors
Factors are the r-objects which are created using a vector. It stores the vector along with the distinct values of the elements in the vector as labels. The labels are always character irrespective of whether it is numeric or character or Boolean etc. in the input vector. They are useful in statistical modeling.

Factors are created using the factor() function.The nlevels functions gives the count of levels.

```{r,}
# Create a vector.
apple_colors <- c('green','green','yellow','red','red','red','green')

# Create a factor object.
factor_apple <- factor(apple_colors)

# Print the factor.
print(factor_apple)
print(nlevels(factor_apple))
```

<font color ="red"><b> Lets try a few functions:</font> </b>
<pre>
relevel()
length()
as.character()
</pre>


### Data Frames
Data frames are tabular data objects. Unlike a matrix in data frame <b>each column can contain different modes of data</b>. The first column can be numeric while the second column can be character and third column can be logical. It is a list of vectors of equal length.

Data Frames are created using the data.frame() function.

```{r,}
# Create the data frame.
BMI <- 	data.frame(
   gender = c("Male", "Male","Female"), 
   height = c(152, 171.5, 165), 
   weight = c(81,93, 78),
   Age = c(42,38,26)
)
print(BMI)


```

<font color ="red"><b> Lets try a few built in functions: </font> </b>
<pre>
dim()
ncol()
nrow()
class()
str()
rownames()
colnames()
</pre>

When you create or read in a dataframe from a file, it always reads strings as factors. You can see this by printing print(BMI$gender). 


### Lists
A list is an R-object which can contain many different types of elements inside it like vectors, functions and even another list inside it.
``` {r, }
# Create a list.
list1 <- list(c(2,5,3),21.3,BMI)

# Print the list.
print(list1)
unlist(list1)
```




## Tibble

Tibbles are a modern take on data frames. They keep the features that have stood the test of time, and drop the features that used to be convenient but are now frustrating (i.e. converting character vectors to factors).

```{r,}
 t<-tibble(x = 1:3, y = list(1:5, 1:10, 1:20))
print(t)
print(t$y)
dim(t)

rownames(t)<-c("3","4","5")

```

## Indexing Vectors, Dataframes and lists

Indexing elements in vectors or dataframe or lists is a very improtant concept.Lets try out indexing in each of these data types.

## Vectors

```{r}

apple_colors[1]

```

## dataframe
```{r}
# To access column 1
BMI[,1]
# To access row 1
BMI[,1]
# Access columns by using $ notation and column names

BMI$gender
# To access a specific element
BMI[2,3]

```

## list

```{r,}
list1
#get the first element of the list which is a vector
list1[[1]]
#get the second element of the first vector in the list
list1[[1]][2]
# get the Age column out of a dataframe which is the 3rd element of the list
list1[[3]][,"Age"]
```
<font color ="red"><b> Lets try to make a list: </font> </b>

Make a list with 2 elements one a vector and second a dataframe.

## Conditional Statements
Decision making is an important part of programming. This can be achieved in R programming using the conditional if...else statement.
### if
The syntax of if statement is:
<pre>
if (test_expression) {
   statement
}
</pre>

```{r}
x <- 5
if(x > 0){
   print("Positive number")
}
```

### if...else statement
The syntax of if...else statement is:
<pre>
if (test_expression) {
   statement1
} else {
   statement2
}
</pre>

The else part is optional and is only evaluated if test_expression is FALSE.

It is important to note that else must be in the same line as the closing braces of the if statement.

```{r}
x<--1
if(x > 0){
   print("Positive number")
} else {
  print("Negative number")
}
```

### Nested if else statements

if ( test_expression1) {
   statement1
} else if ( test_expression2) {
   statement2
} else
   statement4

```{r}

x <- 0
if (x < 0) {
   print("Negative number")
} else if (x > 0) {
   print("Positive number")
} else
   print("Zero")
```

<font color ="red"><b> Exercise: </font> </b>
Write a simple if else statement to check if 5  is an odd number and if it is print("I am odd") else print("I am even")


## Loops
Conceptually, a loop is a way to repeat a sequence of instructions under certain conditions. They allow you to automate parts of your code that are in need of repetition.

The easiest and most frequently used loop in R is <b>for</b> loop. Here is a demonstration of using loops
```{r}
year=c(2015,2016,2017,2018)

for( i in year)
{
  print(i)
  
}


for( i in year)
{
  print(paste0("the year is",i))
  
}

```

<font color ="red"><b> Exercises: </font> </b>

```{r,}
#1. Create a vector 'nums' filled with 30 random normal values.hint(use rnom function)

#2. Write a for loop which goes over the vector nums

# for() {
#3. print the square of each element of nums
#}
```


```{r}

#1. write a for loop that calculate the square of the first 10 elements of nums.

#2. In the for loop print the square value
```


```{r}

# 1. initialize a vector 'square'

#2. write a for loop that calculate the square of the first 10 elements of nums.

#3. Store these values into square vector

#4. Outside of  the for loop print the square value

```

## Functions (optional)

Watch the video 

https://campus.datacamp.com/courses/intermediate-r/chapter-3-functions?ex=1

<font color ="red"><b> Exercise: </font> </b>

Write a function which take two vectors a and b as input and returns a new vector which concatenates the two

## Tidyr functions
tidyr is new package that makes it easy to “tidy” your data. Tidy data is data that’s easy to work with: it’s easy to munge (with dplyr), visualise (with ggplot2 or ggvis) and model (with R’s hundreds of modelling packages). The two most important properties of tidy data are:

Each column is a variable.

Each row is an observation.

Arranging your data in this way makes it easier to work with because you have a consistent way of referring to variables (as column names) and observations (as row indices). When use tidy data and tidy tools, you spend less time worrying about how to feed the output from one function into the input of another, and more time answering your questions about the data.

To tidy messy data, you first identify the variables in your dataset, then use the tools provided by tidyr to move them into columns. tidyr provides three main functions for tidying your messy data: gather(), separate() and spread().

gather() takes multiple columns, and gathers them into key-value pairs: it makes “wide” data longer. Other names for gather include melt (reshape2), pivot (spreadsheets) and fold (databases). Here’s an example how you might use gather() on a made-up dataset. In this experiment we’ve given three people two different drugs and recorded their heart rate:



```{r,}
library(tidyr)
library(dplyr)

messy <- data.frame(
  name = c("Wilbur", "Petunia", "Gregory"),
  a = c(67, 80, 64),
  b = c(56, 90, 50)
)
messy
```

We have three variables (name, drug and heartrate), but only name is currently in a column. We use gather() to gather the a and b columns into key-value pairs of drug and heartrate:

<pre>gather(data, key = "key", value = "value", ..., na.rm = FALSE,
  convert = FALSE, factor_key = FALSE)
</pre>

```{r,gather}
messy %>%
  gather(drug, heartrate, a:b)
```
Sometimes two variables are clumped together in one column. separate() allows you to tease them apart (extract() works similarly but uses regexp groups instead of a splitting pattern or position). Take this example from stackoverflow (modified slightly for brevity). We have some measurements of how much time people spend on their phones, measured at two locations (work and home), at two times. Each person has been randomly assigned to either treatment or control.

```{r,}
set.seed(10)
messy <- data.frame(
  id = 1:4,
  trt = sample(rep(c('control', 'treatment'), each = 2)),
  work.T1 = runif(4),
  home.T1 = runif(4),
  work.T2 = runif(4),
  home.T2 = runif(4)
)
head(messy,8)
```

To tidy this data, we first use gather() to turn columns work.T1, home.T1, work.T2 and home.T2 into a key-value pair of key and time. (Only the first eight rows are shown to save space.)

```{r,tidy}
tidier <- messy %>%
  gather(key, time, -id, -trt)
tidier %>% head(8)
```
Next we use separate() to split the key into location and time, using a regular expression to describe the character that separates them.

```{r,separate}
tidy <- tidier %>%
  separate(key, into = c("location", "time"), sep = "\\.")
tidy %>% head(8)
```

The last tool, spread(), takes two columns (a key-value pair) and spreads them in to multiple columns, making “long” data wider. Spread is known by other names in other places: it’s cast in reshape2, unpivot in spreadsheets and unfold in databases. spread() is used when you have variables that form rows instead of columns. You need spread() less frequently than gather() or separate() so to learn more, check out the documentation and the demos.